A generalized definition of multidimensional item response theory parameters

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects terminology ("index" to "parameter")

Up to this point,
we have derived the maximum slope and its location in the latent space
$`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
If we assume that $`r BASIS_EQ`$ or, more generally,
that $`r LS_BASIS`$ is any orthonormal basis, $`r TRANSFORM_MATRIX`$ will be
also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
In such a case the multidimensional item location and discrimination simplify
to the expressions derived by @reckase_difficulty_1985
and @reckase_discriminating_1991.
As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
which bears no meaning besides its pure algebraic purpose,
we shall refer to them as the *agnostic* version of the indices:

@@

Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
$`r INNER_PROD_COV_COND`$;
in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
*covariance-based* version of the indices:

@@

The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
also due to the strict positivity of $`r MDISC_SYM`$,
are equal to the signs of $`r DISCR_VECTOR`$.
This property leads directly to generalizing the indices to the monotonically
non-increasing case: The measurement direction relative to dimension
$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
When $`r IRF_ABBR`$ is constant with respect to
variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
However, note that this does not necessarily imply that the vector is
parallel to any other axis or strictly contained in the hyperplane formed
by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
orthogonality assumption, but it will depend on the correlation matrix
$`r CORR_MATRIX`$ for the other two versions of the index.

@@

Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
that allows us to represent the test space structure
and the items by transforming their coordinates to
a rectangular Cartesian system.
However, to facilitate the interpretation,
we want to avoid changing the scaling of the parameters.
We will thus prefer the correlation-based over the covariance-based indices.
Thus, we define $`r TRANSFORM_MATRIX`$ such that
$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based index has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

>>>>

Up to this point,
we have derived the maximum slope and its location in the latent space
$`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
If we assume that $`r BASIS_EQ`$ or, more generally,
that $`r LS_BASIS`$ is any orthonormal basis, $`r TRANSFORM_MATRIX`$ will be
also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
In such a case the multidimensional item location and discrimination simplify
to the expressions derived by @reckase_difficulty_1985
and @reckase_discriminating_1991.
As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
which bears no meaning besides its pure algebraic purpose,
we shall refer to them as the *agnostic* version of the parameters:

@@

Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
$`r INNER_PROD_COV_COND`$;
in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
*covariance-based* version of the parameters:

@@

The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
also due to the strict positivity of $`r MDISC_SYM`$,
are equal to the signs of $`r DISCR_VECTOR`$.
This property leads directly to generalizing the parameters to the monotonically
non-increasing case: The measurement direction relative to dimension
$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
When $`r IRF_ABBR`$ is constant with respect to
variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
However, note that this does not necessarily imply that the vector is
parallel to any other axis or strictly contained in the hyperplane formed
by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
orthogonality assumption, but it will depend on the correlation matrix
$`r CORR_MATRIX`$ for the other two versions of the parameter.

@@

Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
that allows us to represent the test space structure
and the items by transforming their coordinates to
a rectangular Cartesian system.
However, to facilitate the interpretation,
we want to avoid changing the scaling of the parameters.
We will thus prefer the correlation-based over the covariance-based parameters.
Thus, we define $`r TRANSFORM_MATRIX`$ such that
$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.
